---
title: "Multiple Imputation via MICE"
author: "Yuchen Ding, Angelina Lu"
date: "2024-10"
output: html_document
---

```{r setup, include=FALSE, tidy=TRUE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE,
                      autodep = TRUE, cache.comments = FALSE,
                      message = FALSE, warning = FALSE,
                      fig.width = 4.5, fig.height = 3)
```

## Overview

This notebook implements **Multiple Imputation by Chained Equations (MICE)**
on a 2024 U.S. presidential election survey dataset with simulated MNAR
(Missing Not At Random) missingness patterns.

Three data conditions are compared:

| Dataset       | Description                                      |
|---------------|--------------------------------------------------|
| `df_self_mnar` | Self-reported missingness pattern (MNAR)        |
| `df_cross_mnar`| Cross-item missingness pattern (MNAR)           |
| `df_cleaned`   | Complete-case baseline                          |

> **Note:** Original survey data are not included in this repository
> (restricted access). The scripts below are designed to run on
> analogous data with the same structure. See `data/generate_demo_data.R`
> to create a compatible synthetic dataset.

---

## 1. Load Data

```{r load-data}
# Replace paths with your own data location
# Demo data can be generated via data/generate_demo_data.R
df_self_mnar  <- read.csv("data/df_self_mnar.csv",  na.strings = "")
df_cross_mnar <- read.csv("data/df_cross_mnar.csv", na.strings = "")
df_cleaned    <- read.csv("data/df_cleaned.csv",    na.strings = "")
```

---

## 2. Explore Missingness

```{r explore-missing}
library(VIM)

missing_counts_self_mnar  <- sort(colSums(is.na(df_self_mnar)),  decreasing = TRUE)
missing_counts_cross_mnar <- sort(colSums(is.na(df_cross_mnar)), decreasing = TRUE)
missing_counts_cleaned    <- sort(colSums(is.na(df_cleaned)),    decreasing = TRUE)

print(missing_counts_self_mnar)
print(missing_counts_cross_mnar)
print(missing_counts_cleaned)
```

```{r aggr-plots}
aggr(df_self_mnar,  col = c("navyblue", "red"), numbers = TRUE, sortVars = TRUE,
     main = "Missing Pattern: Self MNAR")
aggr(df_cross_mnar, col = c("navyblue", "red"), numbers = TRUE, sortVars = TRUE,
     main = "Missing Pattern: Cross MNAR")
aggr(df_cleaned,    col = c("navyblue", "red"), numbers = TRUE, sortVars = TRUE,
     main = "Missing Pattern: Cleaned")
```

---

## 3. MICE Imputation (PMM)

Predictive Mean Matching (`pmm`) is used with `m = 5` imputed datasets
and `maxit = 5` iterations.

```{r mice-impute}
library(mice)

imp_self_mnar  <- mice(df_self_mnar,  m = 5, method = "pmm", maxit = 5, seed = 123)
imp_cross_mnar <- mice(df_cross_mnar, m = 5, method = "pmm", maxit = 5, seed = 123)
imp_cleaned    <- mice(df_cleaned,    m = 5, method = "pmm", maxit = 5, seed = 123)
```

```{r mice-summary}
summary(imp_self_mnar)
summary(imp_cross_mnar)
summary(imp_cleaned)
```

---

## 4. Export Completed Datasets

```{r export}
library(dplyr)

df_self_mnar_MI  <- complete(imp_self_mnar,  include = TRUE)
df_cross_mnar_MI <- complete(imp_cross_mnar, include = TRUE)
df_cleaned_MI    <- complete(imp_cleaned,    include = TRUE)

write.csv(df_self_mnar_MI,  "data/output/df_self_mnar_MI.csv",  row.names = FALSE)
write.csv(df_cross_mnar_MI, "data/output/df_cross_mnar_MI.csv", row.names = FALSE)
write.csv(df_cleaned_MI,    "data/output/df_cleaned_MI.csv",    row.names = FALSE)
```

---

## 5. Visual Comparison: MICE vs KNN

```{r compare-plot}
library(ggplot2)

mi_self_mnar  <- complete(imp_self_mnar)
# knn_self_mnar loaded from 02_knn_imputation.Rmd output

# Example: compare distribution of a target variable
# Replace `target_var` with your variable of interest
# ggplot(mi_self_mnar, aes(x = target_var)) +
#   geom_density(fill = "blue", alpha = 0.4) +
#   labs(title = "Density of MICE Imputation", x = "Values", y = "Density")
```
