---
title: "KNN Imputation with Optimal K Selection"
author: "Yuchen Ding"
date: "2024-11"
output: html_document
---

```{r setup, include=FALSE, tidy=TRUE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE,
                      autodep = TRUE, cache.comments = FALSE,
                      message = FALSE, warning = FALSE,
                      fig.width = 4.5, fig.height = 3)
```

## Overview

This notebook implements **KNN imputation** on three survey question blocks
and selects the optimal `k` by comparing imputation accuracy on artificially
introduced MNAR missingness.

Each question block corresponds to a distinct survey module (e.g., candidate
evaluation, policy preferences). Variable names have been anonymized; the
original data are not included due to access restrictions.

> **Note:** To run this notebook, generate demo data via
> `data/generate_demo_data.R` or supply analogous CSV files.

---

## 1. Load Data

```{r load-data}
library(dplyr)
library(VIM)
library(caret)
library(ggplot2)

# Three survey question blocks (complete + MNAR versions)
# Replace with your own data paths
df_block1_mnar <- read.csv("data/df_block1_mnar.csv", na.strings = "")
df_block2_mnar <- read.csv("data/df_block2_mnar.csv", na.strings = "")
df_block3_mnar <- read.csv("data/df_block3_mnar.csv", na.strings = "")

df_block1 <- read.csv("data/df_block1.csv", na.strings = "")
df_block2 <- read.csv("data/df_block2.csv", na.strings = "")
df_block3 <- read.csv("data/df_block3.csv", na.strings = "")
```

---

## 2. Explore Missingness Patterns

```{r aggr-plots}
aggr(df_block1_mnar, col = c("navyblue", "red"), numbers = TRUE, sortVars = TRUE,
     main = "Missing Pattern: Block 1 MNAR")
aggr(df_block2_mnar, col = c("navyblue", "red"), numbers = TRUE, sortVars = TRUE,
     main = "Missing Pattern: Block 2 MNAR")
aggr(df_block3_mnar, col = c("navyblue", "red"), numbers = TRUE, sortVars = TRUE,
     main = "Missing Pattern: Block 3 MNAR")
```

---

## 3. Define Variable Sets per Block

```{r define-vars}
# Adapt these to match your actual variable names
block1_vars <- paste0("item_", 1:8)   # 8-item block
block2_vars <- paste0("item_", 1:12)  # 12-item block
block3_vars <- paste0("item_", 1:9)   # 9-item block

df_block1      <- df_block1      %>% select(all_of(block1_vars))
df_block1_mnar <- df_block1_mnar %>% select(all_of(block1_vars))

df_block2      <- df_block2      %>% select(all_of(block2_vars))
df_block2_mnar <- df_block2_mnar %>% select(all_of(block2_vars))

df_block3      <- df_block3      %>% select(all_of(block3_vars))
df_block3_mnar <- df_block3_mnar %>% select(all_of(block3_vars))
```

---

## 4. Accuracy Evaluation Function

Compares imputed values against original values *only* at positions
that were missing in the MNAR dataset.

```{r eval-function}
evaluate_imputation_accuracy <- function(imputed, original, original_mnar) {
  vars    <- names(original)
  acc_vec <- sapply(vars, function(col) {
    missing_mask <- is.na(original_mnar[[col]])
    mean(imputed[[col]][missing_mask] == original[[col]][missing_mask],
         na.rm = TRUE)
  })
  mean(acc_vec, na.rm = TRUE)
}
```

---

## 5. Select Optimal K (k = 1 to 10)

```{r find-best-k}
find_best_k_accuracy <- function(data_complete, data_mnar, k_values = 1:10) {
  data_complete[] <- lapply(data_complete, as.factor)
  data_mnar[]     <- lapply(data_mnar,     as.factor)

  accuracy_k <- numeric(length(k_values))

  for (i in seq_along(k_values)) {
    k       <- k_values[i]
    imputed <- kNN(data_mnar, k = k, imp_var = FALSE)
    acc     <- evaluate_imputation_accuracy(imputed, data_complete, data_mnar)
    accuracy_k[i] <- acc
    cat(sprintf("k = %2d  →  Accuracy: %.4f\n", k, acc))
  }

  best_k <- k_values[which.max(accuracy_k)]
  list(best_k = best_k, accuracy_by_k = accuracy_k)
}

set.seed(42)

result_block1 <- find_best_k_accuracy(df_block1, df_block1_mnar)
result_block2 <- find_best_k_accuracy(df_block2, df_block2_mnar)
result_block3 <- find_best_k_accuracy(df_block3, df_block3_mnar)

cat("Best k — Block 1:", result_block1$best_k, "\n")
cat("Best k — Block 2:", result_block2$best_k, "\n")
cat("Best k — Block 3:", result_block3$best_k, "\n")
```

**Results on original data (for reference):**

| Block | Best k | Accuracy range (k=1–10) |
|-------|--------|------------------------|
| Block 1 | 5 | 95.7% – 96.0% |
| Block 2 | 1 | 94.7% – 94.9% |
| Block 3 | 5 | 95.4% – 95.5% |

---

## 6. Apply Best K and Export

```{r apply-best-k}
df_block1_knn <- kNN(df_block1_mnar, k = result_block1$best_k, imp_var = FALSE)
df_block2_knn <- kNN(df_block2_mnar, k = result_block2$best_k, imp_var = FALSE)
df_block3_knn <- kNN(df_block3_mnar, k = result_block3$best_k, imp_var = FALSE)

# Verify no remaining NAs
cat("NAs remaining — Block 1:", sum(is.na(df_block1_knn)), "\n")
cat("NAs remaining — Block 2:", sum(is.na(df_block2_knn)), "\n")
cat("NAs remaining — Block 3:", sum(is.na(df_block3_knn)), "\n")

write.csv(df_block1_knn, "data/output/df_block1_knn.csv", row.names = FALSE)
write.csv(df_block2_knn, "data/output/df_block2_knn.csv", row.names = FALSE)
write.csv(df_block3_knn, "data/output/df_block3_knn.csv", row.names = FALSE)
```
